---
title: "World Cup Data Analysis"
author: "CS"
date: "2022-11-03"
output: 
  beamer_presentation:
    theme: "CambridgeUS"
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(ggplot2)
library(psych)
library(sos)
library(kableExtra)
library(knitr)
library(ggthemes)
wc_dat <- read.csv('matches.csv', header=TRUE)
```

## Introduction

* Data analysis on World cup data from 1930 - 2018
* Based all matches that occurred
* Found via \url{https://www.kaggle.com/datasets/abecklas/fifa-world-cup}
* Contains over 500 records and 15 columns of data

## Data description

```{r, Intro, fig.dim = c(2,2)}
#getting structure of the dataset - based on two observations as makes it simple to display
str(head(wc_dat, 2))
```


## Questions for the analysis

1. What is the trend of total goals scored over every world cup?
2. What is the trend of average goals scored per match over every world cup?
3. Which nations have won the most world cups?
4. Which nations have won the world cup on home soil?
5. Who are the top 6 nations with the most world cup match wins?
6. Who are the top 6 nations with the best average goals per match in the world cup?
7. Who are the biggest world cup losers of all time?


## 1. What is the trend of total goals over every world cup?

::: columns
:::: column
```{r, Q1, echo=FALSE, fig.width =4, fig.height=3}
#grabbing the total goals score for home teams grouped by year
home_scores <- wc_dat %>%
  select(year, home_team, home_score) %>% 
  group_by(year,home_team) %>% 
  summarise(total = sum(home_score), .groups = 'drop')

#doing the same as above but for away
away_scores <- wc_dat %>% 
  select(year, away_team, away_score) %>% 
  group_by(year,away_team) %>% 
  summarise(total = sum(away_score), .groups = 'drop')

#using left join to combine both home and away totals per year
scores <-left_join(home_scores, away_scores, by=c('year'='year', 'home_team'='away_team'))

#now removing nulls and getting the total row sums between the home and away and storing within within a total goals
scores <- scores %>% 
  replace(is.na(.), 0) %>% 
  select(everything()) %>% 
  mutate(total_goals = rowSums(across(c(total.x, total.y))))
colnames(scores) <- c('year', 'nation', 'home_goals', 'away_goals', 'total_goals')

## selecting the year and total goals and grouping by years
yearly_goals <- scores %>% 
  select(year,total_goals) %>% 
  group_by(year) %>% 
  summarise(total_goals = sum(total_goals), .groups = 'drop')

#visualizing our solution where we colour based on year with scale_fill_hue(). Displays total goals each year trend
ggplot(yearly_goals, aes(year, total_goals, fill=as.character(year)))+geom_bar(stat='identity')+
  geom_text(aes(label=year), size=2)+ coord_flip()+scale_fill_hue(h = c(180, 300))+theme(legend.position='none')+labs(x="Goal Total", y="Year")

```

::::
:::: column
* Total goals have increased overtime
* 2014 had the most number of goals scored with 171
* There is a gap in between 1938-1950. Possibly due to WWII (1939-45)
::::
:::

## 2. What is the trend of average goals scored per match over every world cup?
::: columns
:::: column
* The average goals per match is decreasing overtime
* Peaking at 2.7 in 1954
* Some world cups had an average goals per match 1.1
* Stayed fairly consistent/similar since 1962 and onwards
::::
:::: column
```{r, Q2, echo=FALSE, fig.width =4, fig.height=3}
#getting the average goals per home nation grouped by year
home_average <- wc_dat %>%
  select(year, home_score) %>% 
  group_by(year) %>% 
  summarise(homeavg_goals = mean(home_score))
#getting the average goals per away nation grouped by year
away_average <- wc_dat %>%
  select(year, away_score) %>% 
  group_by(year) %>% 
  summarise(awayavg_goals = mean(away_score))
#combining home and away together with left join
average_goals <-left_join(home_average, away_average, by=c('year'='year'))
#simiarly to q1 getting the rowMeans (instead of rowSums) of home and away to get the total average
overall_avg <- average_goals %>% 
  select(everything()) %>% 
  mutate(average_goals = rowMeans(across(c(homeavg_goals, awayavg_goals))))
#producing this result via a gg line plot
ggplot(overall_avg, aes(year, average_goals))+geom_line(colour = "firebrick", size=2)+geom_point(size=2)+labs(x= "Year", y="Average goals")

```
::::
:::

## 3. Which nations have won the most world cups?
::: columns
:::: column
```{r, Q3, echo=FALSE}
#teams who have won the most finals
#fairly simple, filter the stage to finals for winner field, count then arrange in descending to get highest first
winners <- wc_dat %>% 
  select(winning_team,stage) %>% 
  filter(stage=='Final') %>% 
  count(winning_team) %>% 
  arrange(desc(n)) %>% 
  rename(Nation=winning_team, Wins=n)
#storing within a kable table
kable(winners) %>% 
  kable_styling(latex_options = "striped")
```
::::
:::: column
* Brazil is #1!
* Not far behind is Italy with 4
* France the recent winners have 2
* Germany have a total of 4 (if we include both West Germany)
::::
:::

## 4. Which nations have won the world cup on home soil?

::: columns
:::: column
* All have only won once
* 5/9 teams have won the world cup on home turf
* Most recent winner on home turf was France in 1998
::::
:::: column

```{r, Q4, echo=FALSE}
#similar to Q4 but we also filter by if country and winning team are the same result
home_soil <-wc_dat %>% 
  select(winning_team, country, stage, year) %>% 
  filter(stage == 'Final' ) %>% 
  filter(country == winning_team) %>% 
  select(winning_team, year) %>% 
  rename(Nation=winning_team, Year=year)
#display as a kable table
kable(home_soil) %>% 
  kable_styling(latex_options = "striped")
```
::::
:::


## 5. Who are the top 6 nations with the most world cup match wins?
::: columns
:::: column
```{r, Q5, echo=FALSE}
#counting the total for each winning team and arranging in descending to get the top
winning_team <- wc_dat %>%
  select(winning_team) %>% 
  count(winning_team) %>% 
  arrange(desc(n)) %>% 
  na.omit() %>% 
  #use top_n() to filter the top 6
  top_n(6, n) %>% 
  rename(Total_wins= n, Nation=winning_team)
#display the results via kable
kable(winning_team) %>% 
  kable_styling(latex_options = "striped")


```
::::
:::: column
* Brazil dominates
* Germany really have 71
* Brazil and Germany dominate in wins
::::
:::

## 6. Who are the top 6 nations with the best best average goals per match in the world cup?
::: columns
:::: column
- Hungary are #1!
- Turkey, Yugoslavia, Hungary all have done well
- We assume this is due to them having less matches
::::
:::: column
```{r, Q6, echo=FALSE}
#collecting the home average score based on home team
home_average <- wc_dat %>%
  select(home_team, home_score) %>% 
  group_by(home_team) %>% 
  summarise(homeavg_goals = mean(home_score))

#collecting the away average score based on away team
away_average <- wc_dat %>%
  select(away_team, away_score) %>% 
  group_by(away_team) %>% 
  summarise(awayavg_goals = mean(away_score))
#combining home and away via left join
average_goals <-left_join(home_average, away_average, by=c('home_team'='away_team'))
#getting getting the row means of both home and away to get the final average goals per each nation
# arranged in descending as we want the top
overall_avg <- average_goals %>% 
  select(everything()) %>% 
  mutate(average_goals = rowMeans(across(c(homeavg_goals, awayavg_goals)))) %>% 
  select(home_team, average_goals) %>% 
  arrange(desc(average_goals)) %>% 
  #want top 6 so use top_n()
  top_n(6, average_goals) %>% 
  mutate(average_goals=round(average_goals,2)) %>% 
  rename(Nation=home_team, Average_goals = average_goals)
#display as a kable table
kable(overall_avg) %>% 
  kable_styling(latex_options = "striped")
```
::::
:::


## 7. Who are the biggest world cup losers of all time?

::: columns
::: column
```{r, Q7, echo=FALSE}
#this result may be the smallest but there is a lot that need to be considered to procduce this correctly

#firstly grab all nations that have lost a match that does not existing within the winning field
#then getting one of that nation
countries <- wc_dat %>%
  select(winning_team, losing_team, year, outcome) %>% 
  filter(!losing_team %in% winning_team) %>% 
  select(losing_team) %>% 
  distinct() %>% 
  rename(nation=losing_team)

#we also need to consider draws as they are not considered a win or loss
# can be found through the outcome field as 'D' so filter all draws
draw <- wc_dat %>% 
  select(home_team, away_team, outcome) %>% 
  filter(outcome == 'D')

#Now we get the losing teams so we can get the total count
losing_team <- wc_dat %>%
  select(losing_team) %>% 
  count(losing_team) %>% 
  arrange(desc(n)) %>% 
  na.omit() %>% 
  rename(nation = losing_team)

#getting the total score for both home and away to prove that our result is zero
home_sum <- wc_dat %>%
  select(home_team, home_score) %>% 
  group_by(home_team) %>% 
  summarise(home_goals = sum(home_score))
away_sum <- wc_dat %>%
  select(away_team, away_score) %>% 
  group_by(away_team) %>% 
  summarise(away_goals = sum(away_score))
#and combining like previous questions have
total_goals <-left_join(home_sum, away_sum, by=c('home_team'='away_team'))
#now collecting the total amount of goals for both home and away
overall_sum <- total_goals %>% 
  select(everything()) %>% 
  mutate(total_goals = rowSums(across(c(home_goals, away_goals)))) %>% 
  select(home_team, total_goals) %>% 
  arrange(desc(total_goals))

#now we have all our checking methods so we will filter through overall_sum
# we will filter if total goals is zero
# inner join with countries and losing team records
# filter to check if the nation is contained within the home draw field and the away field, it should not
# then our final results will be all teams with no wins and no goals plus their total matches played
biggest_losers <- overall_sum %>% 
  rename(nation=home_team) %>% 
  filter(total_goals == 0) %>% 
  inner_join(countries, by='nation') %>% 
  inner_join(losing_team, by='nation') %>% 
  filter(!nation %in% draw$home_team) %>% 
  filter(!nation %in% draw$away_team) %>% 
  rename(Nation=nation, Matches=n, Goals = total_goals)
#displaying as a kable table
kable(biggest_losers) %>% 
  kable_styling(latex_options = "striped")
```
::::
:::: column
* A 'Biggest loser' is a nation with no wins nor goals
* Canada and China are fairly big nations yet have had poor success
* Canada will be attending  Qatar 2022
::::
:::

## Thank you
- There is still plenty of other things to consider!!!